An Approach to Off-line Handwriting Recognition through Segmentation Techniques and Deep Neural Networks

### 2017-12-17 Report/paper.tex
correction of errors in section 2

\section{Character Recognition}
\label{sec:charrec}
The smallest distinguishable token in a text is a character. These handwritten characters are uniquely
written by each individual and variations even occur when written by the same individual. These variations include the inclination of characters, i.e., the slant, orientation and size.
A way of dealing with these includes removing variations from the input image using preprocessing techniques.
We take another approach, instead of adding rules to minimize variations we made a deep neural net which can recognise any character with a certain probability. The architecture of this deep neural network is described in Section \ref{sec:dnn}.
In order to make the deep neural network recognise characters in images with these variations, we augmented the dataset to include a bigger variety of images. We present these augmentation techniques in Section \ref{sec:preproc}.
The dataset of handwritten characters is described in Section \ref{sec:data}.

\subsection{Dataset}
\label{sec:data}
Before exposing the neural network to new characters, it needs to be familiar with many training samples to be able to classify any new character correctly.
This is a why a good dataset is very important. A dataset should have many examples, much variation, and an even balance between classes. A dataset with these properties
allow the neural network to train effectively and recognise any new character given to the network.

@@

\subsubsection{Normalizing data}
\label{par:norm}
Theoretically feeding normalized data to the neural network returns the same output as before.
In reality however, it is better to normalize data to avoid getting stuck in local optima and to increase training speed of the model \cite{NormGoal}.
We normalize the data by reducing the input dimensions to (64,64,1). A simple threshold is added to improve the contrast in our image.
This allows the foreground to be more distinguishable from the background and allows the neural network to train and recognise more easily.

\subsubsection{Data augmentation}
\label{par:aug}
Data augmentation serves two purposes. It makes the dataset more robust to different kinds of handwriting as we add variations of the original data to the input data.
The second purpose is to increase the size of our dataset. As our dataset only contains 3410 images with handwritten characters, we need more data for the neural network to be able to function properly.
Next we will describe the data augmentation techniques we used. For these techniques we based us on \cite{DataAug}. A more thorough explanation can be found there.

\paragraph{Translations of data}
The dataset is also extended with variations in padding of characters in images. For this purpose we multiply every coordinate of the original image with the following transformation matrix:
\begin{equation}
        \begin{bmatrix}
                1 & 0 & t_x \\
                0 & 1 & t_y
        \end{bmatrix}
\end{equation}
where $t_x$ is the horizontal shift and $t_y$ is the vertical shift. The values for chosen for $t_x$ are -8, 0 and 8, the values chosen for $t_y$ are -4, 0 and 4. All combinations are used except $t_x=0$ and $t_y=0$. The translation only occurs if it does not displace the character outside of the image.
\paragraph{Rotations of data}
Rotations of the image are added to the dataset to make the network more robust to handwritings with different orientations. This matrix is equal to:
\begin{equation}
       \begin{bmatrix}
               \alpha & \beta & (1-\alpha)*center.x - \beta*center.y \\
               -\beta & \alpha & \beta*center.x + (1-\alpha)*center.y
       \end{bmatrix}
\end{equation}
where $\alpha = cos(angle)$ and $\beta = sin(angle)$. $center$ is the coordinate in our image around which is rotated and the angle describes the amount of degrees to rotate in the clockwise direction. We use -30 and 30 as angle.
\paragraph{Scaling of data}
With the purpose of making the network more robust to different sizes of handwriting, we add different scalings of the data. We scale in both the horizontal and vertical direction. The following transformation matrix allows us to scale the input image:
\begin{equation}
       \begin{bmatrix}
               s_x & 0 & 0  \\
               0 & s_y & 0
       \end{bmatrix}
\end{equation}
where $s_x$ is the horizontal scaling factor and $s_y$ is the vertical scaling factor.
We use combinations of values in $[0.75, 1, 1.25, 1.5]$ for $s_x$ and $s_y$ depending on the original size of the character.
\paragraph{Shearing of data}
In order to deal with different kinds of slants we add sheared versions of the image to the dataset.
Shearing displaces each point in a fixed direction, by an amount proportional to its signed distance from a line that is parallel to that direction \cite{Shear}. We used the following shear mapping to achieve this:
\begin{equation}
        \begin{bmatrix}
                1 & s & 0 \\
                0 & 1 & 0
        \end{bmatrix}
\end{equation}
\paragraph{Erosion of data}
The erosion operation of images adds thinner versions of characters to the dataset. Erosion uses a kernel to convolute an image.
The image is scanned using the kernel. The maximum pixel value of the overlapping part between the image and the kernel is computed. This pixel value of the anchor point of the kernel is replaced with the maximum pixel value.
This causes bright regions to get thinner and dark regions to get bigger in the image.
All of the previously discussed preprocessing techniques are also applied to the eroded image. The final dataset contains translated, orientated, sheared and scaled versions of both the original image and the eroded image. A visualitation of the augmentation of an augmented picture can be seen in Figure \ref{fig:augmented}.

\begin{figure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/original}
  \caption{Original}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/translated}
  \caption{Translated}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/rotated}
  \caption{Rotated}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/scaled}
  \caption{Scaled}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/sheared}
  \caption{Sheared}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/eroded}
  \caption{Eroded}
\end{subfigure}
\caption{An image transformed by the discussed augmentation techniques.}
\label{fig:augmented}
\end{figure}
\subsection{Deep neural network}
\label{sec:dnn}
% TODO: make a drawing, see http://www.ais.uni-bonn.de/deep_learning/images/Convolutional_NN.jpg
Humans have millions of connected neurons in the visual cortices which are capable of image processing.
We can recognise most characters without any thought because of this complex neural network inside our brain.
For computers, character recognition is a more difficult task. We could try to write rule-based systems and define how every character should look. This solution is not flexible at all and is very hard to define.
We use a neural network, inspired by how the brain processes images. These deep neural networks contain several layers of connected perceptrons, where each connection is defined by weight and each perceptron is defined by a bias.
The perceptrons take an input and calculate an output using an activation function. This allows perceptrons to make decisions.
Normalized images can be fed to the neural network, which then propagate through several layers and eventually produce as output the probability that the input image has a certain class, one of the possible characters.
Now follows a detailed overview of our deep neural network. This overview is visualized in Figure \ref{fig:dnn}. A detailed explanation on how deep neural networks function is explained by \cite{nnbook}.

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{images/dnn}
  \caption{Visualization of the deep neural network for character recognition. The first row shows the shape a layer. The second row shows the visualization of that layer, the third row shows information about the layer. 'C' indicates a convolutional layer, 'MP' a max-pooling layer and 'FC' a fully-connected layer.}
  \label{fig:dnn}
\end{figure*}

@@

 The input is a normalized image with one color channel. This is fed to a convolutational layer. Convolutional layers allow us to extract features by scanning the image with filters. These features might be edges, corners or any other type of shape in the image. The extracted features are called a feature map.

@@

 This works because images have a spatial structure: close pixels have shared information. By applying a filter, pixels in the same region will use shared biases and weights, which allows the filter to extract particular features.

@@

There are three fully-connected layers each connected with a dropout layer. Dropout randomly drops units and their connections from the network during training. Dropout greatly reduces overfitting in the fully connected layers. \cite{dropout} We do not use this after the convolutional layers as they are very resistant to overfitting \cite{nnbook}.

@@

 The last layer in the neural network is a softmax layer with a cross-entropy cost function. The softmax layer compute a probability distribution (a set of 62 numbers which sum up to 1) using the 62 output neurons from the previous layer. An element in the probability distribution estimates the probability of the input image to classify as a certain character.
 The cross-entropy cost function uses the actual labels to calculate the error between the actual labels and the predicted labels (the previously discussed probabilities). The closer the cross-entropy function is to zero, the better it is at classifying handwritten characters.

@@

First of all, stochastic gradient descent (SGD) is an iterative method which tries to find minima. This function takes a step size, which is known as the learning rate. SGD can be used to minimize the cross-entropy function iteratively by adjusting the weights.

>>>>

\section{Character Recognition}
\label{sec:charrec}
The smallest distinguishable token in a text is a character. These handwritten characters are uniquely
written by each individual and variations even occur when written by the same individual. These variations include the inclination of characters (i.e. the slant) orientation and size.
A way of dealing with these includes removing variations from the input image using preprocessing techniques.
We take another approach, instead of adding rules to minimize variations we made a deep neural net which can recognise any character with a certain probability. The architecture of this deep neural network is described in Section \ref{sec:dnn}.
In order to make the deep neural network recognize characters in images with these variations, we augmented the dataset to include a bigger variety of images. We present these augmentation techniques in Section \ref{sec:preproc}.
The dataset of handwritten characters is described in Section \ref{sec:data}.

\subsection{Dataset}
\label{sec:data}
Before exposing the neural network to new characters, it needs to be familiar with many training samples to be able to classify any new character correctly.
This is why a good dataset is very important. A dataset should have many examples, much variation, and an even balance between classes. A dataset with these properties
allows the neural network to train effectively and recognize any new character given to the network.

@@

\subsubsection{Normalizing data}
\label{par:norm}
Theoretically feeding normalized data to the neural network returns the same output as before.
In reality however, it is better to normalize data to avoid getting stuck in local optima and to increase training speed of the model \cite{NormGoal}.
We normalize the data by reducing the input dimensions to (64,64,1). A simple threshold is added to improve the contrast in our image.
This allows the foreground to be more distinguishable from the background and allows the neural network to train and recognize characters more easily.

\subsubsection{Data augmentation}
\label{par:aug}
Data augmentation serves two purposes. It makes the dataset more robust to different kinds of handwriting as we add variations of the original data to the training samples.
The second purpose is to increase the size of our dataset. As our dataset only contains 3410 images of handwritten characters, we need more data for the neural network to be able to function properly.
Next we will describe the data augmentation techniques we used. For these techniques we based us on \cite{DataAug}. A more thorough explanation can be found there.

\paragraph{Translations of data}
The dataset is extended with variations in padding of characters in images. For this purpose we multiply every coordinate of the original image with the following transformation matrix:
\begin{equation}
        \begin{bmatrix}
                1 & 0 & t_x \\
                0 & 1 & t_y
        \end{bmatrix}
\end{equation}
where $t_x$ is the horizontal shift and $t_y$ is the vertical shift. The values for chosen for $t_x$ are -16, 0 and 16, the values chosen for $t_y$ are -8, 0 and 8. All combinations are used except $t_x=0$ and $t_y=0$. The translation only occurs if it does not move the character outside of the image.
\paragraph{Rotations of data}
Rotations of the image are added to the dataset to make the network more robust to handwritings with different orientations. The transformation matrix is equal to:
\begin{equation}
       \begin{bmatrix}
               \alpha & \beta & (1-\alpha)*center.x - \beta*center.y \\
               -\beta & \alpha & \beta*center.x + (1-\alpha)*center.y
       \end{bmatrix}
\end{equation}
where $\alpha = cos(angle)$ and $\beta = sin(angle)$. $center$ is the coordinate in our image around which is rotated and the angle describes the amount of degrees to rotate in the clockwise direction. We use -30 and 30 as angle.
\paragraph{Scaling of data}
With the purpose of making the network more robust to different sizes of handwriting, we add different scalings of the data. We scale in both the horizontal and vertical direction. The following transformation matrix allows us to scale the input image:
\begin{equation}
       \begin{bmatrix}
               s_x & 0 & 0  \\
               0 & s_y & 0
       \end{bmatrix}
\end{equation}
where $s_x$ is the horizontal scaling factor and $s_y$ is the vertical scaling factor.
We use combinations of values in $[0.75, 1, 1.25, 1.5]$ for $s_x$ and $s_y$ depending on the original size of the character.
\paragraph{Shearing of data}
In order to deal with different kinds of slants we add sheared versions of the image to the dataset.
Shearing displaces each point in a fixed direction, by an amount proportional to its signed distance from a line that is parallel to that direction \cite{Shear}. We used the following shear mapping to achieve this:
\begin{equation}
        \begin{bmatrix}
                1 & s & 0 \\
                0 & 1 & 0
        \end{bmatrix}
\end{equation}
\paragraph{Erosion of data}
The erosion operation of images adds thinner versions of characters to the dataset. Erosion uses a kernel to convolute an image.
The image is scanned using the kernel. The maximum pixel value of the overlapping part between the image and the kernel is computed. This pixel value of the anchor point of the kernel is replaced with the maximum pixel value.
This causes bright regions to get thinner and dark regions to get bigger in the image.

All of the previously discussed preprocessing techniques are also applied to the eroded image. The final dataset contains translated, orientated, sheared and scaled versions of both the original image and the eroded image. A visualitation of the augmentation of a picture can be seen in Figure \ref{fig:augmented}.

\begin{figure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/original}
  \caption{Original}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/translated}
  \caption{Translated}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/rotated}
  \caption{Rotated}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/scaled}
  \caption{Scaled}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/sheared}
  \caption{Sheared}
\end{subfigure}
\begin{subfigure}{0.15\textwidth}
  \centering
  \includegraphics[width=\linewidth]{images/eroded}
  \caption{Eroded}
\end{subfigure}
\caption{An image transformed by the discussed augmentation techniques.}
\label{fig:augmented}
\end{figure}
\subsection{Deep neural network}
\label{sec:dnn}
Humans have millions of connected neurons in the visual cortices which are capable of image processing.
We can recognize most characters without any thought because of this complex neural network inside our brain.
For computers, character recognition is a more difficult task. We could try to write rule-based systems and define how every character should look. This solution is not flexible at all and is very hard to define.
We use a neural network, inspired by how the brain processes images. These deep neural networks contain several layers of connected perceptrons, where each connection is defined by a weight and each perceptron is defined by a bias.
The perceptrons take an input and calculate an output using an activation function. This allows perceptrons to make decisions.
Normalized images can be fed to the neural network, which then propagate through several layers and eventually produce as output the probability that the input image has a certain class, one of the possible characters.
Now follows a detailed overview of our deep neural network. This overview is visualized in Figure \ref{fig:dnn}. A detailed explanation on how deep neural networks function is explained by \cite{nnbook}.

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{images/dnn}
  \caption{Visualization of the deep neural network for character recognition. The first row shows the shape of a layer. The second row shows the visualization of that layer, the third row shows information about the layer. 'C' indicates a convolutional layer, 'MP' a max-pooling layer and 'FC' a fully-connected layer.}
  \label{fig:dnn}
\end{figure*}

@@

 The input is a normalized image with one color channel. This is fed to a convolutional layer. Convolutional layers allow us to extract features by scanning the image with filters. These features might be edges, corners or any other type of shape in the image. The extracted features are called a feature map.

@@

 This works because images have a spatial structure; i.e. close pixels have shared information. By applying a filter, pixels in the same region will use shared biases and weights, which allows the filter to extract particular features.

@@

There are three fully-connected layers of which the first two are connected with a dropout layer and the last one is the output layer.
Dropout randomly drops units and their connections from the network during training. Dropout greatly reduces overfitting in the fully connected layers. \cite{dropout} We do not use this after the convolutional layers as they are very resistant to overfitting \cite{nnbook}.

@@

 The last layer in the neural network is a softmax layer with a cross-entropy cost function. The softmax layer computes a probability distribution (i.e. a set of 62 numbers which sum up to 1) using the 62 output neurons from the previous layer. An element in the probability distribution estimates the probability of the input image to classify as a certain character.
 The cross-entropy cost function uses the actual labels to calculate the error between the actual labels and the predicted labels (the previously discussed probabilities). The closer the cross-entropy value is to zero, the better the neural network is at classifying handwritten characters.

@@

First of all, stochastic gradient descent (SGD) is an iterative method that tries to find minima. This function takes a step size which is known as the learning rate. SGD can be used to minimize the cross-entropy function iteratively by adjusting the weights.
